Demo: Counterpoint Analysis and Synthesis

### 2022-08-16 doc/farm22/abstract.tex
2nd diagram fixed

The initial prototype \citep{HaskellCounterpoint} of this tool was
written in Haskell, chosen due to the availability of the
high-quality and feature-rich SMT library SBV \citep{SBV}. It was then
ported to the Agda Music Tools library \cite{MusicTools}, using Agda's
Haskell foreign function interface (FFI) to call SBV. There is also an
Agda interface \citep{Schmitty} to SMT and we might convert to using
that at some point in the future.

@@

Here Haydn has supplied the top voice, marked ``C.F.'' for cantus
firmus, and Beethoven's task was to supply the remaining three voices
following first species counterpoint rules. He uses only tones from
triads, all of them complete save the last and all in root position
save the second triad in bar 4. Since the notes form triads they
automatically satisfy consonance rules (only perfect intervals, thirds
or sixths between any pair of notes vertically); note that a perfect
fourth is prohibited in two part strict counterpoint but allowed with
more voices. Perfect fourths can be found between the top two voices
in the first, fourth and last bars.

@@

For simplicity we now focus only on the top two voices. Feeding
Beethoven's notes into the tool as two part first species
counterpoint, it easily finds and reports errors in the use of perfect
fourths, missing octaves at the boundaries, and similar motion into
fifths and octaves. We can disable the boundary rules as they are not
relevant and relax the consonance rule to allow perfect fourths, leaving
only the motion errors. We can replace the note F that Haydn fixed
with a hole, and run the tool in synthesis mode using the same
rules. It makes the same fix Haydn did. Perhaps we gave it too much
guidance, so instead we can create three holes in a row near the
error, which generates the solution shown in
Figure~\ref{fig:b146fix3}. Note that this introduces a series of five
parallel 6ths, and there is a ``soft'' rule that one should not have
long sequences of 3rds or 6ths as that also diminishes independence
of voices. However we can easily add a constraint limiting parallel
intervals to at most two in a row; the solution returned changes the G
in bar three to an A, breaking the chain with a fifth, a small
improvement but one might hope for more.

@@

This can be done by adding a constraint to increase contrary motion
(in which both voices move in opposite directions), and in fact we can
try generating our own two part counterpoint following Haydn's cantus
firmus, fixing only the first and last of Beethoven's notes. To help
create better quality counterpoint we add constraints to limit the
number of leaps (horizontal intervals of more than a major third) to
at most one and require at least six instances of contrary motion,
which turns out to be maximal. Note that it would be extremely
difficult for a human being with no computer assistance to generate
counterpoint following such restrictions, but the SMT solver instantly
returns the solution shown in Figure~\ref{fig:b146gen3}.

>>>>

The initial prototype \citep{HaskellCounterpoint} of this tool was
written in Haskell, chosen due to the availability of the
high-quality and feature-rich SMT library SBV \citep{SBV}. It was then
ported to the Agda Music Tools library \cite{MusicTools}, using Agda's
Haskell foreign function interface (FFI) to call SBV. There is also an
Agda interface \citep{Schmitty} to SMT; we might convert to using
that at some point in the future.

@@

Here Haydn has supplied the top voice, marked ``C.F.'' for cantus
firmus, and Beethoven's task was to supply the remaining three voices
following first species counterpoint rules. He uses only tones from
triads, all of them complete save the last and all in root position
save the second triad in bar 4. Since the notes form triads they
automatically satisfy consonance rules (only perfect intervals, thirds
or sixths between any pair of notes vertically); note that a perfect
fourth is prohibited in strict two part counterpoint but allowed with
more voices. Perfect fourths can be found between the top two voices
in the first, fourth and last bars.

@@

For simplicity we now focus only on the top two voices. Feeding
Beethoven's notes into the tool as two part first species
counterpoint, it easily finds and reports errors in the use of perfect
fourths, missing octaves at the boundaries, and similar motion into
fifths and octaves. We can disable the boundary rules as they are not
relevant and relax the consonance rule to allow perfect fourths, leaving
only the motion errors. We can replace the note F that Haydn fixed
with a hole, and run the tool in synthesis mode using the same
rules. It makes the same fix Haydn did. Perhaps we gave it too much
guidance, so instead we can create three holes in a row near the
error, which generates the solution shown in
Figure~\ref{fig:b146fix3}, with the three generated notes inside the
blue box. Although there are several possible solutions, the SMT
solver has chosen to retain the third A while changing the first two
notes to A as well; the only difference with Haydn's correction is
that the first note was changed from G to A as well. Although this
creates a satisfying oblique motion into a perfect fifth between bars
2 and 3, this same motion occurs between bars 5 and 6; furthermore we
now have four A notes in a row in addition to the three G notes at the
end of the second voice. For a middle voice this is not too bad, but
viewed as just two voices it is a big dull.

@@

This can be fixed by adding a constraint to limit the number of repeated
notes, but we can also try generating our own two part counterpoint
following Haydn's cantus firmus, fixing only the first and last of
Beethoven's notes. To help create better quality counterpoint we add
constraints to limit the number of leaps (horizontal intervals of more
than a major third) to at most one and require at least six instances
of contrary motion, which turns out to be maximal. Note that it would
be extremely difficult for a human being with no computer assistance
to generate counterpoint following such restrictions, but the SMT
solver instantly returns the solution shown in
Figure~\ref{fig:b146gen3}.
